Uncertainty Estimation Methods for Countering Attacks on Machine-Generated Text Detectors

### 2025-05-16 paper/main.tex
some changes

\begin{abstract}
	In this work, we explore the use of uncertainty estimation methods to enhance the quality of machine-generated text detectors against various attacks, such as homoglyphs, paraphrasing and noise injection. These attacks are not only used to bypass detection but also to test the resilience of detectors. We will test the hypothesis that uncertainty estimation can provide a more sustainable approach, eliminating the need for constant retraining. The research will evaluate this hypothesis in two scenarios: when only the text is available and when access to the model's internal states is also provided. The planned experiments aim to validate the results and compare them with current state-of-the-art solutions.
\end{abstract}

@@

We formally define our detection framework through a \textbf{classification model} \( F: \mathcal{D} \rightarrow \mathcal{C} \), where \(\mathcal{D}\) represents the space of input texts and \(\mathcal{C} = \{0,1\}\) denotes the binary classification space (0 for human-written texts, 1 for machine-generated outputs). The research will explore two approaches to calculating uncertainty.

@@

\subsection {Black-box methods}   
In many modern models, internal states and architecture are not available for study, but even in this case there are many methods for estimating uncertainty based only on the response classification model \(F(x)\). Black-box methods were also used in the work\citep{Polygraph}. The task is to study these methods and compare their effectiveness with the white-box methods.

@@

>>>>

\begin{abstract}
	This study investigates the application of uncertainty estimation methods to enhance the quality of machine-generated text detectors when processing data containing attacks such as homoglyphs, paraphrasing, and noise injection. These attacks are not only used to bypass detection but also serve to test the robustness of detectors. We examine the hypothesis that uncertainty estimation methods can provide a more resilient approach, eliminating the need for continuous retraining across various attack types. We propose a method combining uncertainty estimation with classifiers based on hidden representations of language models. Experiments on the M4GT and RAID datasets demonstrate competitive accuracy (ROC-AUC 0.8977) with significantly lower computational costs compared to fine-tuning large language models (LLMs).
\end{abstract}

@@

We formally define our detection framework through a \textbf{classification model} \( F: \mathcal{D} \rightarrow \mathcal{C} \), where \(\mathcal{D}\) represents the space of input texts and \(\mathcal{C} = \{0,1\}\) denotes the binary classification space (0 for human-written texts, 1 for machine-generated outputs). There are two main types of uncertainty estimation methods.

@@

\subsection {Black-box methods}   
In many modern models, internal states and architecture are not available for study, but even in this case there are many methods for estimating uncertainty based only on the response classification model \(F(x)\). However, these methods typically provide significantly less information than white-box approaches, and thus their application will not be considered in this work.

@@

In Figure 1, you can see a UMAP plot of our texts from RAID categorized by attack types. It is noticeable that only the homoglyph attack forms a distinct cluster, while the other types of attacks cannot be reliably distinguished from each other based solely on uncertainty metrics.

\begin{figure}[ht]
  \centering
  \includegraphics[width=1\linewidth]{umap.jpg}
  \caption{
    UMAP-projection of texts from RAID categorized by attack types.   }
  \label{fig:umap}
\end{figure}
